Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2014-06-04 paper.tex
minor

For the case of carrier amplitude the situation is more complicated.  A simulation study conducted by Pauluzzi~\cite{Pauluzzi2000} suggests that known non-data-aided estimators for carrier amplitude exhibit bias.  The bias is often small when the noise variance is small, but typically becomes significant as the noise variance increases.  Neverthess, Cram\'{e}r-Rao bounds have been derived for \emph{unbiased} estimators of carrier amplitude~\cite{Alagha_crb_snr_bpsk_qpsk_2001,Delmas_exact_crb_psk_2008}.  Unfortunately, these bounds may not be meaningful since known non-data-aided amplitude estimators appear to be biased~\cite{Pauluzzi2000}.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cram\'{e}r-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convincing response is that, while the specific results given here apply only to the least squares estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection}~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010,McKilliam_pps_unwrapping_tsp_2014}, would potentially allow this.  
More ambitiously, our techniques might allow analysis of non-data-aided channel estimators for modern communications systems such as multiple antenna systems, systems employing orthogonal frequency division multiplexing, and systems that must withstand rapidly time varying channels~\cite{4138046,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cram\'{e}r-Rao lower bounds.

@@

We will have particular use of complex random variables of the form $1 + W$ where $W$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + W.
\]
The joint pdf of $R$ and $\Phi$ can be shown to be
 \begin{equation}\label{eq:pdfRPhi}
 f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{2\pi\sqrt{r^2 - 2r\cos\phi + 1}}.
 \end{equation}
Since $\cos\phi$ has period $2\pi$ and is even on $[-\pi,\pi]$ it follows that $f(r,\phi)$ has period $2\pi$ and is even on $[-\pi,\pi]$ with respect to $\phi$.  The mean of $R e^{j\Phi}$ is equal to one because the mean of $W$ is zero.  So,
%\begin{equation}\label{eq:expectRepartRphi}
\[
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\]
%\end{equation}
where $\Re(\cdot)$ denotes the real part, and
\begin{equation}\label{eq:expectImpartRphi}
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\end{equation}
where $\Im(\cdot)$ denotes the imaginary part, and $\expect$ denotes the expected value.  Whenever there is no chance of ambiguity we write simply $\expect X$ rather than $\expect(X)$ to denote the expected value of the random variable $X$.  For two random variables $X$ and $Y$ the notation $\expect X Y$ will always mean the expected value of the product $XY$, that is, $\expect(XY)$ and not $\expect(X)Y$.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the sample mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilot symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality}.   %(and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  
The prediction is made by dividing the asymptotic covariance matrix by $L$.  The theorem accurately predicts the behaviour of the phase estimator when SNR is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Our asymptotic theory cannot be expected to model the behaviour of the estimator for fixed $L$ as SNR goes to zero.  Similar conclusions are made in the literature~\cite{ViterbiViterbi_phase_est_1983,6365847}.  As the SNR increases the variance of the estimators converge to that of the estimator where all symbols are pilots, i.e. $\abs{P} = L$.  As expected, the figures indicate that the accuracy of the estimator increases with the number of pilot symbols.  One interesting observation is that, at low SNR, the slope of the MSE curve when no pilot symbols exist is steeper than the slope when pilot symbols do exist.

>>>>

For the case of carrier amplitude the situation is more complicated.  A simulation study conducted by Pauluzzi~\cite{Pauluzzi2000} suggests that known non-data-aided estimators for carrier amplitude exhibit bias.  The bias is often small when the noise variance is small, but typically becomes significant as the noise variance increases.  Nevertheless, Cram\'{e}r-Rao bounds have been derived for \emph{unbiased} estimators of carrier amplitude~\cite{Alagha_crb_snr_bpsk_qpsk_2001,Delmas_exact_crb_psk_2008}.  Unfortunately, these bounds may not be meaningful since known non-data-aided amplitude estimators appear to be biased~\cite{Pauluzzi2000}.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cram\'{e}r-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convincing response is that, while the specific results given here apply only to the least squares estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection}~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straightforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010,McKilliam_pps_unwrapping_tsp_2014}, would potentially allow this.  
More ambitiously, our techniques might allow analysis of non-data-aided channel estimators for modern communications systems such as multiple antenna systems, systems employing orthogonal frequency division multiplexing, and systems that must withstand rapidly time varying channels~\cite{4138046,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cram\'{e}r-Rao lower bounds.

@@

We will have particular use of complex random variables of the form $1 + W$ where $W$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + W.
\]
The joint pdf of $R$ and $\Phi$ can be shown to be
 \begin{equation}\label{eq:pdfRPhi}
 f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{2\pi\sqrt{r^2 - 2r\cos\phi + 1}}.
 \end{equation}
Since $\cos\phi$ has period $2\pi$ and is even on $[-\pi,\pi]$ it follows that $f(r,\phi)$ has period $2\pi$ and is even on $[-\pi,\pi]$ with respect to $\phi$.  The mean of $R e^{j\Phi}$ is equal to one because the mean of $W$ is zero.  So,
%\begin{equation}\label{eq:expectRepartRphi}
\[
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\]
%\end{equation}
where $\Re(\cdot)$ denotes the real part, and
\begin{equation}\label{eq:expectImpartRphi}
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\end{equation}
where $\Im(\cdot)$ denotes the imaginary part, and $\expect$ denotes the expected value.  Whenever there is no chance of ambiguity we write simply $\expect X$ rather than $\expect(X)$ to denote the expected value of a random variable $X$.  For two random variables $X$ and $Y$ the notation $\expect X Y$ will always mean the expected value of the product $XY$, that is, $\expect(XY)$ and not $\expect(X)Y$.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the sample mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilot symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality}.   %(and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  
The prediction is made by dividing the asymptotic covariance matrix by $L$.  The theorem accurately predicts the behaviour of the phase estimator when SNR is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Our asymptotic theory cannot be expected to model the behaviour of the estimator for fixed $L$ as SNR goes to zero.  Similar observations have been made in the literature~\cite{ViterbiViterbi_phase_est_1983,6365847}.  As the SNR increases the variance of the estimators converge to that of the estimator where all symbols are pilots, i.e. $\abs{P} = L$.  As expected, the figures indicate that the accuracy of the estimator increases with the number of pilot symbols.  One interesting observation is that, for low SNR, the slope of the MSE curve when no pilot symbols exist is steeper than the slope when pilot symbols do exist.  This indicates that pilot symbols are particularly important when the SNR is low.

### 2014-06-04 response2.tex
minor

\item \textbf{Comment:} ``the data symbols are not known to the receiver and must also be estimated'' $\to$ I would drop the second part of the sentence because (e.g. Viterbi and Vterbi estimator) non-data-aided estimators typically do not actually estimate the unknown data symbols \\
\textbf{Response:} Done.

@@

\item \textbf{Comment:} p. 4: ``systems such as mulitple anntannae'' $\to$ should be ``multiple antennae'' \\
\textbf{Response:} Reviewer 1 comment~\eqref{eq:rev1:multantenna} wanted ``antenna'' and we have used this spelling. ``Multiple'' is now spelt correctly.

@@

``Whenever there is no chance of ambiguity we write simply $\expect X$ rather than $\expect(X)$ to denote the expected value of the random variable $X$.  For two random variables $X$ and $Y$ the notation $\expect X Y$ will always mean the expected value of the product $XY$, that is, $\expect(XY)$ and not $\expect(X)Y$.''

@@

\item \textbf{Comment:} although different figures are shown for M=2,4,8 (phase estimation), the effect of the constellation size is not addressed in the text  \\
\textbf{Response:}  We have included the follows sentences to Section~\ref{sec:simulations}:

``As expected, the figures indicate that the accuracy of the estimator increases with the number of pilot symbols.  One interesting observation is that, at low SNR, the slope of the MSE curve when no pilot symbols exist is steeper than the slope when pilot symbols do exist.''

@@

``As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Our asymptotic theory cannot be expected to model the behaviour of the estimator for fixed $L$ as SNR goes to zero.  Similar conclusions are made in the literature~\cite{ViterbiViterbi_phase_est_1983,6365847}.''

@@

\textbf{Response:} Figure~\ref{fig:biasvarcomp} in this document displays such a plot.  The number of symbols is set to $L=4096$.  The solid line shows that asymptotic squared bias 
\[
\rho_0^2\big(1 - G(0)\big)^2
\]
and the dashed line in the asymptotic variance
\[
\rho_0^2\frac{p B_1 + d B_2}{L}.
\]
The number of pilots consider is $\abs{P}=0$ and $\abs{P}=L/2$.  The obvious relationship is observed: the bias dominates the variance at low SNR and the variance dominates the bias at high SNR.  We believe that this plot is not particularly informative and have not included it in the manuscript.

@@

\item \textbf{Comment:}  There are typografical errors in the words  ``multiple antenna'' in line 2, page 2, section 1 \\
\textbf{Response:} Fixed.

>>>>

\item \textbf{Comment:} ``the data symbols are not known to the receiver and must also be estimated'' $\to$ I would drop the second part of the sentence because (e.g. Viterbi and Viterbi estimator) non-data-aided estimators typically do not actually estimate the unknown data symbols \\
\textbf{Response:} Done.

@@

\item \textbf{Comment:} p. 4: ``systems such as mulitple anntannae'' $\to$ should be ``multiple antennae'' \\
\textbf{Response:} Reviewer 1 comment~\eqref{eq:rev1:multantenna} wanted ``antenna'' and we have used this spelling. ``Multiple'' is now spelled correctly.

@@

``Whenever there is no chance of ambiguity we write simply $\expect X$ rather than $\expect(X)$ to denote the expected value of a random variable $X$.  For two random variables $X$ and $Y$ the notation $\expect X Y$ will always mean the expected value of the product $XY$, that is, $\expect(XY)$ and not $\expect(X)Y$.''

@@

\item \textbf{Comment:} although different figures are shown for M=2,4,8 (phase estimation), the effect of the constellation size is not addressed in the text  \\
\textbf{Response:}  We have included the following sentences to Section~\ref{sec:simulations}:

``As expected, the figures indicate that the accuracy of the estimator increases with the number of pilot symbols.  One interesting observation is that, for low SNR, the slope of the MSE curve when no pilot symbols exist is steeper than the slope when pilot symbols do exist.  This indicates that pilot symbols are particularly important when the SNR is low.''

@@

``As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$, and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Our asymptotic theory cannot be expected to model the behaviour of the estimator for fixed $L$ as SNR goes to zero.  Similar observations have been made in the literature~\cite{ViterbiViterbi_phase_est_1983,6365847}.''

@@

\textbf{Response:} Figure~\ref{fig:biasvarcomp} in this document displays such a plot.  The number of symbols is set to $L=4096$.  The solid line shows that asymptotic squared bias 
\[
\rho_0^2\big(1 - G(0)\big)^2
\]
and the dashed line is the asymptotic variance
\[
\rho_0^2\frac{p B_1 + d B_2}{L}.
\]
The number of pilots considered is $\abs{P}=0$ and $\abs{P}=L/2$.  The obvious relationship is observed: the bias dominates the variance at low SNR and the variance dominates the bias at high SNR.  We believe that this plot is not particularly informative because the same conclusion can be drawn from Figures~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK}.  For this reason we have not included this plot in the manuscript.

@@

\item \textbf{Comment:}  There are typographical errors in the words  ``multiple antenna'' in line 2, page 2, section 1 \\
\textbf{Response:} Fixed.
